### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
done

Ecologists' intense interest in drawing inferences about species
interactions ---especially competition---from presence-absence data has
a long history (MacArthur 1958, Diamond 1975, Connor et al. 2013). If
nontrophic species interactions are important drivers of community
assembly patterns, then we should expect to see their influence in our
data sets. Despite decades of work and several major controversies,
however (Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). More generally, it
can be difficult to reason about the complex web of direct and indirect
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal and environmental filtering (Connor et al. 2013). For this
reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force. For example, even though the two shrub
species in Figure 1 compete with one another for resources at a
mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
If this sort of effect is common, then significant deviations from
independence will not---by themselves---provide convincing evidence of
species' direct effects on one another.

@@

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov
networks, which generalize partial correlations to non-Gaussian data,
have been used in many scientific fields to model associations between
various kinds of ``particles''. For example, a well-studied network
called the Ising model has played an important role in our understanding
of physics (where nearby particles tend to align magnetically with one
another; Cipra 1987). In spatial contexts, these models have been used
to describe interactions between adjacent grid cells (Harris 1974,
Gelfand et al. 2005). In neurobiology, they have helped researchers
determine which neurons are connected to one another by modeling the
structure in their firing patterns (Schneidman et al. 2006). Following
recent work by Azaele et al. (2010) and Fort (2013), I suggest that
ecologists could similarly treat species as the interacting particles in
this modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously imposed
severe approximations that produced unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
it is computationally feasible to fit these models exactly, the approach
has become worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov
networks work and how they can be used to simulate landscape-level data
and to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

Here, \(\alpha_{i}\) is the amount that the presence of species \(i\)
contributes to the log-probability of \(\vec{y}\); it directly controls
the prevalence of species \(i\). Similarly, \(\beta_{ij}\) is the amount
that the co-occurrence of species \(i\) and species \(j\) contributes to
the log-probability, and controls how often the two species will be
found together (Figure 2A, Figure 2B). \(\beta\) thus acts as an analog
of the partial covariance, but for non-Gaussian networks. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce
assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). Calculating
\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the
billions. Numerous techniques are available for working with Markov
networks that keep the computations tractable, either through analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below (a Markov network, two
correlation-based methods and a null modeling approach) was then
evaluated on its ability to correctly infer that the two shrub species
competed with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix B).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

@@

The first two interaction measures were the simple and partial
covariances between each pair of species' data vectors on the landscape
(Albrecht and Gotelli 2001). Because partial covariances are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on ridge regression {[}Wieringen and Peeters
(2014); i.e.~linear regression with a Gaussian prior{]}. For these
analyses, I set the ridge parameter to 0.2 divided by the number of
sites on the landscape.

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retains the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\)-statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} For the first simulated landscape
(three species), I kept the evaluation simple and qualitative: any
method that reliably determined that the two shrub species were
negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. For each estimate of a species pair's interactions, I
used this regression to calculate the squared error associated with
method that produced it. Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical
method, and presence/absence of environmental filtering across all 12
replicates; the mean squared errors associated with these subsets of the
data determined the proportion of variance explained by each method
under different conditions.

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations 94\% of the time and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: the Markov network explained 54\% of the variance
in the ``true'' parameters, followed by partial covariances (33\%), and
sample covariances (22\%).

The null model scores initially explained 12\% of the variation. After
manually reducing the value of one especially strong outlier
(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%.
Figure 3 reflects the adjusted version of the results.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational
data, even when direct interactions are largely overwhelmed by indirect
effects (Figure 1) or environmental effects (lower panels of Figure 3).
For cases where fitting a Markov network is infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly accurate approximation. On the other hand, inferences may
be much less reliable in cases where the ``true'' interaction matrices
include guild structure or trophic levels, which could make the
\(\beta\) coefficients much more difficult to infer (particularly for
linear methods; Loh and Wainwright (2013)). In these cases, it may be
necessary to include prior information about the nature of species'
interactions from natural history, experiments, or trait data. The
rosalia package (Harris 2015a) has built-in mechanisms for incorporating
this kind of information, if it can be expressed as a prior probability
distribution on \(\beta\) or a penalty on the likelihood.

Ecologists will also need experiments and natural history knowledge to
pin down the exact nature of the interactions identified by the network
(e.g.~whether or not negative interactions are consumptive, or which
species in a positively-associated pair is facilitating the other),
particularly when real pairs of species can reciprocally influence one
another's occurrence probabilities in multiple ways simultaneously; the
\(\beta\) coefficients in Markov networks have to reduce this complexity
to a single number. In short, partial correlations and Markov networks
both prevent us from mistaking marginal associations for conditional
ones, but they can't tell us the underlying biological mechanisms at
work.

Despite these limitations, Markov networks have enormous potential to
improve ecological inferences. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Strong et al. 1984). Equation 1 can be used
to calculate the expected prevalence of a species in the absence of
biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and Hastie 2012).
Competition's effect on prevalence in a Markov network can thus be
calculated by subtracting this value and the observed prevalence (cf
Figure 2D).

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications, from spatial statistics
(Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006) to models
of human behavior (Lee et al. 2013). Modeling species interactions using
the same framework would thus allow ecologists to tap into an enormous
set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states (phase transitions).

This modeling approach is also highly extensible, even when it is
inconvenient to compute the likelihood exactly. For example, the mistnet
software package for joint species distribution modeling (Harris 2015b)
can fit \emph{approximate} Markov networks to large species assemblages
(\textgreater{}300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple
ecological processes into a common model could help ecologists to
disentangle different factors that can confound simpler co-occurrence
analyses (cf Connor et al. 2013). Numerous other extensions are
possible: similar networks can be fit with a mix of discrete and
continuous variables, for example (Lee and Hastie 2012). There are even
methods (Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that
would allow the coefficient linking two species in an interaction matrix
to vary as a function of the abiotic environment or of third-party
species that tip the balance between facilitation and exploitation
(Bruno et al. 2003).

>>>>

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009), and thus important
ecological processes remain poorly understood. More generally, it can be
difficult to reason about the complex web of direct and indirect species
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal and environmental filtering (Connor et al. 2013). For this
reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur, versus the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force. For example, even though the two shrub
species in Figure 1 compete with one another for resources at a
mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
If this sort of effect is common, then significant deviations from
independence will not generally provide convincing evidence of species'
direct effects on one another.

@@

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov
networks, which generalize partial correlations to non-Gaussian data
(Lee and Hastie 2012, Loh and Wainwright 2013), have been used in many
scientific fields for decades to model associations between various
kinds of ``particles''. For example, a well-studied network named the
Ising model has played an important role in our understanding of physics
(where nearby particles tend to align magnetically with one another;
Cipra 1987). In spatial contexts, these models have been used to
describe interactions between adjacent grid cells (Harris 1974, Gelfand
et al. 2005). In neurobiology, they have helped researchers determine
which neurons are connected to one another by modeling the structure in
their firing patterns (Schneidman et al. 2006). Following recent work by
Azaele et al. (2010) and Fort (2013), I suggest that ecologists could
similarly treat species as the interacting particles in the same
modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously imposed
severe approximations that produced unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
it is computationally feasible to fit these models exactly, the approach
has become worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov
networks work and how they can be used to simulate landscape-level data
or to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the probability that the two species will be found together
(Figure 2A, Figure 2B). \(\beta\) thus acts as an analog of the partial
covariance, but for non-Gaussian networks. Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce
assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). Calculating
\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species, the number of bookkeeping operations
required for exact inference spirals exponentially into the billions and
beyond. Numerous techniques are available for working with Markov
networks that keep the computations tractable, e.g.~via analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below (a Markov network, two
correlation-based methods and a null model) was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate replicate landscapes
with varying numbers of species and sites (Appendix B). For half of the
simulated landscapes, I treated each species' \(\alpha\) coefficient as
a constant, as described above. For the other half, I treated the
\(\alpha\) coefficients as linear functions of two abiotic environmental
factors that varied from location to location across the landscape
(Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise biotic interactions from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with
most analyses of presence-absence data for co-occurrence patterns---the
inference procedure did not have access to any information about the
environmental or spatial variables that helped shape the landscape (cf
Connor et al. 2013, Blois et al. 2014).

@@

The first two interaction measures were the sample covariances and the
partial covariances between each pair of species' data vectors on the
landscape (Albrecht and Gotelli 2001). Because partial covariances are
undefined for landscapes with perfectly-correlated species pairs, I used
a regularized estimate based on ridge regression {[}Wieringen and
Peeters (2014); i.e.~linear regression with a Gaussian prior{]}. For
these analyses, I set the ridge parameter to 0.2 divided by the number
of sites on the landscape.

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retains the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\) statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} I evaluated each method qualitatively
on the simulated landscapes based on Figure 1: any method that reliably
determined that the two shrub species were negatively associated passed;
other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. For each method, I calculated the squared deviations
between each species pair's rescaled estimate and the ``true''
interactions used to generate the assemblages. Finally, I averaged these
squared errors for each combination of species richness, landscape size,
statistical method, and presence/absence of environmental filtering
across all 12 replicates; the mean squared errors associated with these
subsets of the data determined the proportion of variance explained by
each method under different conditions.

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations on 94\% of simulated landscapes, and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: overall, the Markov network explained 54\% of the
variance in the ``true'' parameters, followed by partial covariances
(33\%), and sample covariances (22\%).

The null model scores initially explained only 12\% of the variation.
After manually reducing the value of one especially strong outlier
(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%
(Appendix B). Figure 3 reflects the adjusted version of the results.

@@

The results presented above show that Markov networks, unlike a common
null modeling approach, can recover species' pairwise interactions from
observational data. The Markov networks were successful even when direct
interactions are largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where
fitting a Markov network is computationally infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly useful approximation. The success of the partial
correlation's success on simulated data may not carry over to real data
sets, however; Loh and Wainwright (2013) suggest that the linear
approximations may be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels). On the
other hand, if ecologists are familiar enough with the natural history
of their study systems to describe this kind of structure as a prior
distribution on the parameters or as a penalty on the likelihood, then
real-world results might be even better than those shown in Figure 3.

Ecologists will also need natural history to pin down the exact nature
of the interactions identified by a network model (e.g.~which species in
a positively-associated pair is facilitating the other), particularly
when real pairs of species can reciprocally influence one another in
multiple ways simultaneously (Bruno et al. 2003); the \(\beta\)
coefficients in Markov networks have to reduce this complexity to a
single number. In short, partial correlations and Markov networks both
help prevent us from mistaking marginal associations for conditional
ones, but they can't tell us the underlying biological mechanisms at
work.

Despite these limitations, Markov networks have enormous potential to
improve ecological inferences. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
Equation 1 can be used to calculate the expected prevalence of a species
in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value and the observed
prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications, from spatial statistics
(Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006) to models
of human behavior (Lee et al. 2013). Modeling species interactions using
the same framework would thus allow ecologists to tap into an enormous
set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states (i.e.~phase
transitions; Cipra (1987)).

This modeling approach is also highly extensible, even when it is
inconvenient to compute the likelihood exactly. For example, the mistnet
software package for joint species distribution modeling (Harris 2015b)
can fit \emph{approximate} Markov networks to large species assemblages
(\textgreater{}300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple
ecological processes into a common model could help ecologists to
disentangle different factors that can confound simpler co-occurrence
analyses (cf Connor et al. 2013). Numerous other extensions are
possible: Markov networks can be fit with a mix of discrete and
continuous variables, for example (Lee and Hastie 2012). There are even
methods (Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that
would allow the coefficient linking two species in an interaction matrix
to vary as a function of the abiotic environment or of third-party
species that tip the balance between facilitation and exploitation
(Bruno et al. 2003).
